Towards a Machine Learning approach for Aggregate Computing

### 2021-11-25 paper-2021-woa-ac-ml.tex
fix acronyms (Aggregate Computing => AC, Multi-Agent Reinforcement Learning => MARL, Reinforcement Learning => RL, Supervised Learning => SL

\newcommand{\unibo}{\textsc{Alma Mater Studiorum} -- \textit{Università di Bologna}}
\newcommand{\disi}{\textit{Computer Science and Engineering - DISI}}
\newcommand{\scafi}{{ScaFi}}
\newcommand{\scalaloci}{{ScalaLoci}}
\newcommand{\scafiloci}{{ScaFiLoci}}
\newcommand{\scalainline}[1]{\mintinline[fontsize=\small]{scala}{#1}}
%%%%% LEGENDA FOR NOTES %%%%%
%% RED: constraints
%% YELLOW: suggestions
%% CYAN: todos
%% GREEN: outline
\newcommand{\constraints}[1]{\todo[inline, color=red]{#1}}
\newcommand{\suggestions}[1]{\todo[inline, color=yellow]{#1}}
\newcommand{\todos}[1]{\todo[inline, color=cyan]{\textbf{TODO}: #1}}
\newcommand{\outline}[1]{\todo[inline, color=green]{#1}}
%% Utility commands
\newcommand{\hybridaggregate}{\textit{Hybrid Aggregate Computing}}
\newcommand{\round}{\texttt{round}}
\newcommand{\export}{\texttt{export}}
\sloppy
\begin{document}

@@

In this direction, Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}(AC) is an innovative paradigm, according to which
 developers can express collective self-organising behaviours through a functional program specification.
%
Indeed, an aggregate program (i.e. a program written with Aggregate Computing) consists of the manipulation of \textit{computational fields}, a distributed
 data structure where each point in the spacetime is associated with the result of a computation
 performed by a node.
%
%These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
% This calculus describes a set of minimal constructs to express any spatiotemporal computation.

@@

In particular, in this paper,
 we investigate how Aggregate Computing can be extended with Reinforcement Learning (RL) through a case study,
 where we outline the performance improvement of the ``hop count" program --- 
 i.e. an aggregate program that produces a field that stores the distance from a source zone. 
 Furthermore, we show that the learnt policy scales up to different system sizes.

@@

In this work, we focus on Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}, a global-to-local approach where
 the self-organising collective behaviour is expressed in terms of manipulation of \textit{computational fields}.
The latter could be seen as the extension of electromagnetic fields to Computer Science since it consists of a 
 map from any devices to their computational value (i.e. the result of a computation).
The computational field manipulation is ruled by \textit{field calculus}, which define the constructs needed 
 to express spatio-temporal computation~\cite{DBLP:conf/coordination/AudritoBDV18}.
The novelty of Aggregate Computing consists in its composable (functional-like) system-size independent program definition: 
using basic constructs we identify the main patterns (\textit{building-blocks}) that could be reusable at the application level.
By opportunistically crafting these building-blocks it is possible to verify prominent properties such as eventual consistency~\cite{DBLP:journals/taas/BealVPD17} 
 (i.e. the ensemble behaviour is independent of the underlying network details)
 and self-stabilisation~\cite{DBLP:journals/tomacs/ViroliABDP18} (i.e. the ability of a system to recover from arbitrary changes).

Aggregate Computing is studied in various simulated scenarios ranging from smart cities to crowds, 
 thanks to Alchemist~\cite{DBLP:journals/jos/PianiniMV13} (a meta-simulator for pervasive systems) and to programming languages and toolchains that support 
 this paradigm, such as ScaFi~\cite{DBLP:conf/ecoop/CasadeiV16}, FCCP~\cite{DBLP:conf/acsos/Audrito20} and Protelis~\cite{PianiniSAC2015}.

@@

\subsection{Computational Model}
An aggregate program execution consists of a continuously node-point execution of a \round{}, 
 that is the atomical part of aggregate computing. Each node can interact only with-in its neighbourhoods sharing the \export{}, namely the result produced by the evaluation of a \round{}.

@@

In this section, we analyse the characteristics of the aggregate computing paradigm,
 map these to relevant machine learning contributions in the literature,
 and point out different learning perspectives about the integration of Aggregate Computing with Machine Learning

\paragraph{Multi-and many-agent system}
%
An aggregate system is a multi-agent system
 and, often, a \emph{many}-agent system
 where a large number (hundreds or more)
 of autonomous entities are programmed to achieve 
 some collective behaviour through \emph{repeated} 
 sensing, computation, communication, and actuation steps.
%
Due to the high stochasticity of the environment,
 it is almost impossible to know and
 program the optimal behaviour for all agents in advance.
 This uncertainty results in the need of creating intelligent agents
 so that they can learn the optimal behaviour and adapt to environmental changes.
%
In recent decades there has been an emerging trend in the use of Reinforcement Learning (RL) 
 in multi-agent settings (called Multi-Agent Reinforcement Learning), as a powerful, robust and adaptive learning paradigm.
 Progress has been considerable and a wide range of algorithms are now available.

Multi-Agent Reinforcement Learning is a conjunction of Game theory and Reinforcement Learning, 
 and there are several (even orthogonal) viewpoints on which researchers have been focused.

@@

The main problem with most of the solutions available in the literature is that they consider a small number of agents (or at least test them on small games).
 In Aggregate Computing instead, we need to find a technique that does not depend on the system node count.

@@

\subsection{Integrating Learning Mechanisms into Aggregate Computing}
We now outline the different perspectives about Aggregate Computing with learning, 
 to clarify the current and the future research path.
%
\paragraph{Application} Learning can be useful at different levels in our framework.
 First of all, learning can be applied to the building blocks definition.
% 
Indeed, most of our results and applications are based on them.
%
The building block improvement may consist either of learning a correction factor or by entirely learning the behaviour.

@@

\paragraph{Learning problem} 
The learning problem in multi-agent settings can be posed as \textit{Independent} and \textit{Team} learning. 
In the foster, each agent perform the learning process and while in the latter,
 the learning process is performed only by one agent and then shared with the collective (in a typical CDTE setting).
% 
Team learning is usually used in the ``swarm'' system, due to the homogeneous behaviour 
 --- i.e. each agent performs the same program. 
 Here, a common reward signal exists and evaluate the overall collective behaviour.
% 
In Aggregate Computing, each agent follows the \textit{same} aggregate program shared within the entire system, so Team learning can 
 be applied seamlessly.

On the other hand, Independent Learning is usually applied in Collective
 (Self-)Adaptive System \cite{csas-and-marl}, a field near to Aggregate Computing.

@@

\paragraph{Learning technique}
Aside from what kind of problem we set up, we can then use different Machine Learning techniques,
 such as Reinforcement Learning or Supervised Learning.
%
Supervised learning is rare in CSAS -- or in the ``swarm" like system in general.
%
Novel works~\cite{DBLP:conf/corl/TolstayaGPP0R19} leverage Graph Neural network to learn how agents should communicate, 
 supposing to know the right behaviour (i.e. produced via global vision simulation or by emulation of natural phenomena).

In Aggregate Computing, Supervised Learning can be applied when we want to build up new building blocks from zero. 
% 
Indeed, we eventually know the right state of the system, but we do not know how the system can reach it.

@@

\paragraph{Learning time}
Finally, learning can be performed \textit{offline}, \textit{online} or \textit{offline an then online} (\textit{mixed time}).
% 
The foster is the common scenario in multi-agent settings.
 Here indeed, we can leverage simulation to know the right computational field and then use one of the novels CTDE techniques. 
Mixed-time learning can be a mid-level complexity direction instead.
 Here, simulation can always be exploited to reach a good behaviour, but at runtime, 
 the behaviour can be adapted based on the new environmental situations. 
Finally, we can imagine that learning can be performed online. 
 It is a common situation where we do not know the environment and also simulation can be very hard to accomplish. 
 In mixed and online cases, Reinforcement Learning is the most suitable model, because agents \textit{learn by doing}.

\section{Aggregate Computing with Reinforcement}\label{aggregate-and-rl}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{img/algorithm-learning.pdf}
  \caption{This figure shows the main steps of RL in Aggregate Computing.
  Each agent starts with his copy of a Q table. 
  Then, during the simulation, they refine them Q tables changing q values according to their experience (different colours mean different values). 
  In the end, a new $Q_{max}$ table will be created aggregating each agent table. 
  Finally, a new episode starts using this $Q_{max}$ as a starting Q table.
  }
  \label{fig:aggregate-q-learning}
\end{figure}
As first experiments, we leverage RL to improve building blocks. 
% 
In this section, we briefly describe how Reinforcement Learning can be injected into the computation loop,
 and what Reinforcement Learning we used with Aggregate Computing.

@@

Speaking of Reinforcement Learning settings, we need to define the state function and the reward function.
%
The execution needs to be distributed, so the state will be a function from the local node context:
$$
K := \Omega(\textit{Context})
$$

@@

As a first step towards a complete integration between Aggregate Computing and RL, we decide to set up an Independent learning technique, 
 by which each agent try to improve our current local policy (shown in \Cref{fig:aggregate-q-learning}). 
% 
We made this choice inspired by other work in CSAS~\cite{csas-and-marl} and swarm systems~\cite{nguyen2018swarm}.

@@

A naive implementation of this kind of problem can be described in Aggregate Computing as:
\begin{minted}[fontsize=\scriptsize]{scala}
def hopCount(): Int = {
  rep(Double.PositiveInfinity) { 
    hopCounts => mux(source) 
    { 0 } 
    { minHoodPlus(nbr{hopCounts}) + 1 }
  }
}
\end{minted}
However, this solution suffers from the slow-rising problem. 
 It consists of a slow convergence time when an equilibrium situation is broken.
So, our first intuition is to accelerate convergence through learning.
 In particular, the previous solution can be rewritten as:

@@

\section{Discussion and Concluding Remarks}\label{conclusion}
In this article, we have discussed a preliminary example of how Aggregate Computing can be integrated with Reinforcement Learning. 
We have then compared Aggregate Computing problems with current Machine Learning solutions in MASs systems.
%
We have defined a possible roadmap towards full support of learning mechanisms in Aggregate Computing. %
As a proof of concept, we have discussed preliminary use case where 
 learning successfully enhances a standard Aggregate Computing program.
Furthermore, we have shown that the learnt program can easily scale up to system size.

>>>>

\newcommand{\unibo}{\textsc{Alma Mater Studiorum} -- \textit{Università di Bologna}}
\newcommand{\disi}{\textit{Computer Science and Engineering - DISI}}
\newcommand{\scafi}{{ScaFi}}
\newcommand{\scalaloci}{{ScalaLoci}}
\newcommand{\scafiloci}{{ScaFiLoci}}
\newcommand{\scalainline}[1]{\mintinline[fontsize=\small]{scala}{#1}}
%%%%% LEGENDA FOR NOTES %%%%%
%% RED: constraints
%% YELLOW: suggestions
%% CYAN: todos
%% GREEN: outline
\newcommand{\constraints}[1]{\todo[inline, color=red]{#1}}
\newcommand{\suggestions}[1]{\todo[inline, color=yellow]{#1}}
\newcommand{\todos}[1]{\todo[inline, color=cyan]{\textbf{TODO}: #1}}
\newcommand{\outline}[1]{\todo[inline, color=green]{#1}}
%% Utility commands
\newcommand{\round}{\texttt{round}}
\newcommand{\export}{\texttt{export}}
\sloppy
\begin{document}

@@

In this direction, Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}(AC) is an innovative paradigm, according to which
 developers can express collective self-organising behaviours through a functional program specification.
%
Indeed, an aggregate program (i.e. a program written with AC) consists of the manipulation of \textit{computational fields}, a distributed
 data structure where each point in the spacetime is associated with the result of a computation
 performed by a node.
%
%These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
% This calculus describes a set of minimal constructs to express any spatiotemporal computation.

@@

In particular, in this paper,
 we investigate how AC can be extended with Reinforcement Learning (RL) through a case study,
 where we outline the performance improvement of the ``hop count" program --- 
 i.e. an aggregate program that produces a field that stores the distance from a source zone. 
 Furthermore, we show that the learnt policy scales up to different system sizes.

@@

In this work, we focus on Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}, a global-to-local approach where
 the self-organising collective behaviour is expressed in terms of manipulation of \textit{computational fields}.
The latter could be seen as the extension of electromagnetic fields to Computer Science since it consists of a 
 map from any devices to their computational value (i.e. the result of a computation).
The computational field manipulation is ruled by \textit{field calculus}, which define the constructs needed 
 to express spatio-temporal computation~\cite{DBLP:conf/coordination/AudritoBDV18}.
The novelty of AC consists in its composable (functional-like) system-size independent program definition: 
using basic constructs we identify the main patterns (\textit{building-blocks}) that could be reusable at the application level.
By opportunistically crafting these building-blocks it is possible to verify prominent properties such as eventual consistency~\cite{DBLP:journals/taas/BealVPD17} 
 (i.e. the ensemble behaviour is independent of the underlying network details)
 and self-stabilisation~\cite{DBLP:journals/tomacs/ViroliABDP18} (i.e. the ability of a system to recover from arbitrary changes).

AC is studied in various simulated scenarios ranging from smart cities to crowds, 
 thanks to Alchemist~\cite{DBLP:journals/jos/PianiniMV13} (a meta-simulator for pervasive systems) and to programming languages and toolchains that support 
 this paradigm, such as ScaFi~\cite{DBLP:conf/ecoop/CasadeiV16}, FCCP~\cite{DBLP:conf/acsos/Audrito20} and Protelis~\cite{PianiniSAC2015}.

@@

\subsection{Computational Model}
An aggregate program execution consists of a continuously node-point execution of a \round{}, 
 that is the atomical part of AC. Each node can interact only with-in its neighbourhoods sharing the \export{}, namely the result produced by the evaluation of a \round{}.

@@

In this section, we analyse the characteristics of the AC paradigm,
 map these to relevant machine learning contributions in the literature,
 and point out different learning perspectives about the integration of AC with Machine Learning

\paragraph{Multi-and many-agent system}
%
An aggregate system is a multi-agent system
 and, often, a \emph{many}-agent system
 where a large number (hundreds or more)
 of autonomous entities are programmed to achieve 
 some collective behaviour through \emph{repeated} 
 sensing, computation, communication, and actuation steps.
%
Due to the high stochasticity of the environment,
 it is almost impossible to know and
 program the optimal behaviour for all agents in advance.
 This uncertainty results in the need of creating intelligent agents
 so that they can learn the optimal behaviour and adapt to environmental changes.
%
In recent decades there has been an emerging trend in the use of RL 
 in multi-agent settings -- called Multi-Agent Reinforcement Learning (MARL) -- as a powerful, robust and adaptive learning paradigm.
 Progress has been considerable and a wide range of algorithms are now available.

MARL is a conjunction of Game theory and RL, 
 and there are several (even orthogonal) viewpoints on which researchers have been focused.

@@

The main problem with most of the solutions available in the literature is that they consider a small number of agents (or at least test them on small games).
 In AC instead, we need to find a technique that does not depend on the system node count.

@@

\subsection{Integrating Learning Mechanisms into Aggregate Computing}
We now outline the different perspectives about AC with learning, 
 to clarify the current and the future research path.
%
\paragraph{Application} Learning can be useful at different levels in our framework.
 First of all, learning can be applied to the building blocks definition.
% 
Indeed, most of our results and applications are based on them.
%
The building block improvement may consist either of learning a correction factor or by entirely learning the behaviour.

@@

\paragraph{Learning problem} 
The learning problem in multi-agent settings can be posed as \textit{Independent} and \textit{Team} learning. 
In the foster, each agent perform the learning process and while in the latter,
 the learning process is performed only by one agent and then shared with the collective (in a typical CDTE setting).
% 
Team learning is usually used in the ``swarm'' system, due to the homogeneous behaviour 
 --- i.e. each agent performs the same program. 
 Here, a common reward signal exists and evaluate the overall collective behaviour.
% 
In AC, each agent follows the \textit{same} aggregate program shared within the entire system, so Team learning can 
 be applied seamlessly.

On the other hand, Independent Learning is usually applied in Collective
 (Self-)Adaptive System \cite{csas-and-marl}.

@@

\paragraph{Learning technique}
Aside from what kind of problem we set up, we can then use different Machine Learning techniques,
 such as RL or Supervised Learning (SL).
%
SL is rare in CSAS -- or in the ``swarm" like system in general.
%
Novel works~\cite{DBLP:conf/corl/TolstayaGPP0R19} leverage Graph Neural network to learn how agents should communicate, 
 supposing to know the right behaviour (i.e. produced via global vision simulation or by emulation of natural phenomena).

In AC, S) can be applied when we want to build up new building blocks from zero. 
% 
Indeed, we eventually know the right state of the system, but we do not know how the system can reach it.

@@

\paragraph{Learning time}
Finally, learning can be performed \textit{offline}, \textit{online} or \textit{offline an then online} (\textit{mixed time}).
% 
The foster is the common scenario in multi-agent settings.
 Here indeed, we can leverage simulation to know the right computational field and then use one of the novels CTDE techniques. 
Mixed-time learning can be a mid-level complexity direction instead.
 Here, simulation can always be exploited to reach a good behaviour, but at runtime, 
 the behaviour can be adapted based on the new environmental situations. 
Finally, we can imagine that learning can be performed online. 
 It is a common situation where we do not know the environment and also simulation can be very hard to accomplish. 
 In mixed and online cases, RL is the most suitable model, because agents \textit{learn by doing}.

\section{Aggregate Computing with Reinforcement}\label{aggregate-and-rl}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{img/algorithm-learning.pdf}
  \caption{This figure shows the main steps of RL in AC.
  Each agent starts with his copy of a Q table. 
  Then, during the simulation, they refine them Q tables changing q values according to their experience (different colours mean different values). 
  In the end, a new $Q_{max}$ table will be created aggregating each agent table. 
  Finally, a new episode starts using this $Q_{max}$ as a starting Q table.
  }
  \label{fig:aggregate-q-learning}
\end{figure}
As first experiments, we leverage RL to improve building blocks. 
% 
In this section, we briefly describe how RL can be injected into the computation loop,
 and what RL we used with AC.

@@

Speaking of RL settings, we need to define the state function and the reward function.
%
The execution needs to be distributed, so the state will be a function from the local node context:
$$
K := \Omega(\textit{Context})
$$

@@

As a first step towards a complete integration between AC and RL, we decide to set up an Independent learning technique, 
 by which each agent try to improve our current local policy (shown in \Cref{fig:aggregate-q-learning}). 
% 
We made this choice inspired by other work in CSAS~\cite{csas-and-marl} and swarm systems~\cite{nguyen2018swarm}.

@@

A naive implementation of this kind of problem can be described in AC as:
\begin{minted}[fontsize=\scriptsize]{scala}
def hopCount(): Int = {
  rep(Double.PositiveInfinity) { 
    hopCounts => mux(source) 
    { 0 } 
    { minHoodPlus(nbr{hopCounts}) + 1 }
  }
}
\end{minted}
However, this solution suffers from the slow-rising problem. 
 It consists of a slow convergence time when an equilibrium situation is broken.
So, our first intuition is to accelerate convergence through learning.
 In particular, the previous solution can be rewritten as:

@@

\section{Discussion and Concluding Remarks}\label{conclusion}
In this article, we have discussed a preliminary example of how AC can be integrated with RL. 
We have then compared AC problems with current Machine Learning solutions in MASs systems.
%
We have defined a possible roadmap towards full support of learning mechanisms in AC. %
As a proof of concept, we have discussed preliminary use case where 
 learning successfully enhances a standard AC program.
Furthermore, we have shown that the learnt program can easily scale up to system size.
